---
title: "Multi-Omics Showcase: Bulk RNA-seq Analysis"
author: "Sara Azidane"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: cosmo
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(DESeq2)
library(airway)
library(ggplot2)
library(plotly)
library(DT)
library(dplyr)
```

# Showcase Overview <span style="color:#1f77b4;"></span>

This project showcases a **simple bulk RNA-seq analysis** as I would do it using R and inluding  **QC, normalization, differential expression, interactive visualization (which for me are super important to share results), and interactive tables (also critical for results discussion)**. 

# Data <span style="color:#ff7f0e;"></span>

For this showcase, I will be using the **Airway dataset** from Bioconductor: human airway smooth muscle cells, treated with dexamethasone vs untreated.

```{r load-data}
data(airway)
airway$dex <- relevel(airway$dex, "untrt")

# Build DESeq2 design
dds <- DESeqDataSet(airway, ~ dex)

# QC and filter
keep <- rowSums(counts(dds) >= 10) >= 4
dds <- dds[keep,]

# Run DE
dds <- DESeq(dds)
res <- results(dds)

# Prepare data frame
res_df <- as.data.frame(res) %>%
  dplyr::mutate(
    gene = rownames(.),
    sig = case_when(
      padj < 0.05 & log2FoldChange > 1  ~ "Up",
      padj < 0.05 & log2FoldChange < -1 ~ "Down",
      TRUE ~ "NS"
    ),
    padj = ifelse(is.na(padj), 1, padj)
  )
```

# Interactive Tables <span style="color:#2ca02c;">📝</span>

## DE Results Table

```{r de-table}
DT::datatable(res_df, filter = 'top', options = list(pageLength = 10, scrollX = TRUE))
```

## Summary Table

```{r summary-table}
sum_table <- res_df %>%
  group_by(sig) %>%
  summarise(count = n())

DT::datatable(sum_table, options = list(dom = 't'))
```

# PCA <span style="color:#d62728;"></span>

```{r pca-plot}
vsd <- vst(dds)
pca_data <- plotPCA(vsd, intgroup="dex", returnData=TRUE)

pca_plot <- ggplot(pca_data, aes(PC1, PC2, color=dex, label=name)) +
  geom_point(size=5, alpha=0.6) +
  geom_text(vjust=-1.5, size=3) +
  scale_color_manual(values=c("#1f77b4", "#ff7f0e")) +
  ggtitle("PCA of Bulk RNA-seq Samples") +
  theme_minimal(base_size=14)

pca_plot
```

# Interactive Volcano Plot <span style="color:#9467bd;">🌋</span>

```{r volcano-plot}
volc <- ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj), color = sig, text = gene)) +
  geom_point(alpha = 0.7, size = 3) +
  scale_color_manual(values = c("Up" = "#e41a1c", "Down" = "#377eb8", "NS" = "#4daf4a")) +
  geom_vline(xintercept = c(-1,1), linetype="dashed", color="black") +
  geom_hline(yintercept = -log10(0.05), linetype="dashed", color="black") +
  theme_minimal(base_size=14) +
  ggtitle("Bulk RNA-seq Volcano Plot") +
  xlab("Log2 Fold Change") +
  ylab("-Log10 Adjusted P-value")

# Convert to interactive plotly
volc_plotly <- ggplotly(volc, tooltip = c("text","x","y"))
volc_plotly

```

```{r knit-folder, include=FALSE}
# This is only relevant when knitting interactively
output_file <- "../results/rna_bulk_showcase.html"
```
