---
title: "Single-Cell RNA-seq Showcase"
author: "Sara Azidane"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: cosmo
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(scRNAseq)
library(SingleCellExperiment)
library(scater)
library(scran)
library(ggplot2)
library(plotly)
library(DT)
library(dplyr)
```

# Project Overview <span style="color:#1f77b4;"></span>

This is a **single-cell RNA-seq analysis showcase**, designed to highlight skills in R, visualization, and reproducible reporting. It includes **QC, normalization, dimensionality reduction, clustering, interactive tables, and interactive plots** for portfolio purposes.

# Data <span style="color:#ff7f0e;"></span>

We are using the **Reprocessed Allen Brain Atlas dataset** from the `scRNAseq` Bioconductor package.

```{r load-data}
sce <- ReprocessedAllenData()
# Basic info about the dataset
sce
```

# QC and Filtering <span style="color:#2ca02c;"></span>

```{r qc-filter}
qc <- perCellQCMetrics(sce, assay.type="tophat_counts")
low_lib <- isOutlier(qc$sum, nmads=3, type="lower")
sce <- sce[, !low_lib]
sce
```

# Normalization and HVG Selection <span style="color:#d62728;"></span>

```{r normalization}
sce <- computeSumFactors(sce, assay.type="tophat_counts")
sce <- logNormCounts(sce, assay.type="tophat_counts")
dec <- modelGeneVar(sce, assay.type="tophat_counts")
top_hvgs <- getTopHVGs(dec, n=2000)
top_hvgs[1:10]
```

# PCA and UMAP <span style="color:#9467bd;">️</span>

```{r pca-umap}
sce <- runPCA(sce, subset_row=top_hvgs)
sce <- runUMAP(sce, dimred="PCA")

umap_plot <- plotUMAP(sce, colour_by="Core.Type") + 
  ggtitle("UMAP of Single-Cell RNA-seq") +
  theme_minimal(base_size=14)
umap_plot
```

# Interactive Volcano-Like Plot for Marker Genes <span style="color:#ff69b4;"></span>

```{r markers, eval=FALSE}
# Example marker gene detection
clusters <- quickCluster(sce, method="igraph", assay.type="tophat_counts" )
sce$cluster <- factor(clusters)
markers <- findMarkers(sce, groups = sce$cluster, assay.type = "tophat_counts")

# Take the first cluster as an example
marker_df <- as.data.frame(markers[[1]]) %>%
  mutate(gene = rownames(.))

marker_plot <- ggplot(marker_df, aes(x = logFC.2, y = -log10(FDR), text = gene)) +
  geom_point(size = 3, alpha = 0.7) +
  theme_minimal(base_size = 14) +
  ggtitle("Marker Gene Volcano-Like Plot") +
  scale_color_manual(values = RColorBrewer::brewer.pal(8, "Set2"))

marker_plotly <- ggplotly(marker_plot, tooltip = c("text","x","y"))
marker_plotly



```

# Interactive Tables <span style="color:#2ca02c;"></span>

```{r tables}
DT::datatable(as.data.frame(colData(sce)), options = list(pageLength = 10, scrollX = TRUE))
```


```{r knit-folder, include=FALSE}
# This is only relevant when knitting interactively
output_file <- "../results/scRNAseq_showcase.html"
